### 2025-09-25 paper/sec/evaluation.tex

To evaluate the system we used a sample of the \SciGen dataset~\citep{moosavi21}. We aggregated the generated
expression in the linguistic categories from Table~\ref{tab:natural-language-forms}.
Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and
without target-value sharing. The results show that providing the target value (text fragment or numerical
value) is crucial for achieving a high success rate. Moreover, the \InterpretationAgent generates a correct
expression in approximately XX\% of the cases.

The system achieved performance above 70\% in comparison, data retrieval, and min/max search tasks.
In contrast, accuracy decreased significantly in expressions requiring differences (20\%) and in ranking tasks (50\%).

Figure~\ref{fig:success_rate_by_complexity} reports the success rate as a function of the number of categories assigned to each expression.
Fluid instructions combining multiple operations were associated with multiple categories.
Results show that the success rate was 75\% for simple expressions (one category), increased to 85\% for two categories,
and dropped to 58\% when three categories were involved.

In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
correct expression when no reference string was given.
On the \SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
Without the target string, performance decreased moderately to 39.4\%, indicating that the model retains substantial
ability to reconstruct valid solutions even without explicit guidance

>>>>

To evaluate the system we used a sample of the \SciGen dataset~\citep{moosavi21}. We aggregated the generated
expression in the linguistic categories from Table~\ref{tab:natural-language-forms}.
Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and
without target-value sharing. The results show that providing the target value (text fragment or numerical
value) is crucial for achieving a high success rate. Moreover, the \InterpretationAgent generates a correct
expression in approximately in 74.9\% of the cases (S.D. 3.0\%).

The system achieved performance above 68\% in comparison, 77.3\% in data retrieval, and 97\% in min/max search tasks.
In contrast, accuracy decreased significantly in expressions requiring differences (20\%) and in ranking tasks (0\%).

Figure~\ref{fig:success_rate_by_complexity} reports the success rate as a function of the number of categories assigned to each expression.
Fluid instructions combining multiple operations were associated with multiple categories.
Results show that the success rate was 62\% for simple expressions (one category), increased to 91\% for two categories,
while no correct expression was generated for expression with 3 categories.

In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
correct expression when no reference string was given.
On the \SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
Without the target string, performance decreased moderately to 57.1\%, indicating that the model retains substantial
ability to reconstruct valid solutions even without explicit guidance
